Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-11-12 Dail_Mad_ext.tex
to results

The zero-inflated Poisson distribution can be applied to recruitment and 
population growth terms as well as initial abundance. For example, 
the recruitment term of the constant-recruitment DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{i,t} \sim \left\{
\begin{aligned}
\mathrm{Pois}(0) &\; \text{with probability} \; \psi \\
\mathrm{Pois}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
\begin{comment}
When both initial abundance and dynamics are modeled as
zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case,
$\psi$ can also be
modeled as varying over space). In the latter case, it makes
sense
for the model to "remember" which sites are inside the range of
the
species between the initial abundance stage and each time step
of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the geometric-recruitment, exponential, Ricker-logistic, or Gompertz-logistic
models
(without immigration) should have no effect when initial
abundance is
also zero-inflated unless $\psi$ is time-varying (because these
dynamic
models cannot recover from zero abundance).
\end{comment}
We have implemented
zero-inflated dynamics in the Bayesian framework using program
\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

\subsection{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, even in
its most stripped down form it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\mathrm{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models include demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- through random 
binomial draws for survival and Poisson draws for recruitment.  Our population
growth extensions use a single Poisson draw for abundance: $N_{i,t} \sim 
\mathrm{Pois}(f(N_{i,t-1}))$.  This is also a commonly used (and mathematically equivalent) 
method of modeling demographic stochasticity 
\citep{bonsall_hastings:2004,melbourne_hastings:2008}.

@@

Another way to model environmental stochasticity is as lognormal variation in observed local population 
growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  This could be applied to both density-independent and density-dependent models.
For example, for the Ricker-logistic + immigration model:
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
Three variations suggest themselves.  The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second, 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
for $\nu_{i,t}$ with distance-dependent (geographic or otherwise) correlation coefficients
between sites in a year might be appropriate.

\subsection{Random effects of observers}

@@

Current BBS trend estimators deal with this problem by
treating observer identity as a random %(as opposed to a fixed)
effect \citep{link_sauer:2002,sauer_link:2011}.
%(Link and Sauer 2002, Sauer and Link 2011).
%This allows observer-specific differences in detection probability,
%but assumes that observers are selected at random from a pool
%of potential observers.  Models that contain both random and
%fixed effects are referred to as mixed models.  Often in mixed
%models (as in this case), the random effect is modeled not
%because it of interest in itself, but to avoid bias in the
%estimates of the fixed effects.
%
To include random observer effects in DM models and these
extensions, Eq.~\ref{eq:p1} can be modified to:
\begin{gather}
X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{N}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
where $X_{i,j,t}$ is the number of bird recorded at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
the standard deviation of the random observer effects (also on the logit scale). 
We have implemented random observer effects in the Bayesian
framework using program \textbf{JAGS} \citep[version 3.2.0, see Appendix]{plummer:2003}.
% (version 3.2.0; Plummer 2003)
%with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.

\subsection{Forecasting Future Population Size}

State-space models allow for inference about future population size,
a critical component of population viability analysis (PVA) as well
as conservation planning. Let $N^+_{t}$ denote population size at some
future point in time and $[N^+_t]$ denote the probability
distribution for this random variable. If we can specify this
distribution, we can make statements about the
probability that $N^+_t = k$ where $k$ is any possible
population size including zero.

@@

%Relative to the computational challenges associated with the
%empirical Bayes methods, 
A fully Bayesian approach to PVA is straightforward to
implement with MCMC. Because population size at any future point %Furthermore, b
in time is regarded as a standard, unknown parameter, computing the
posterior distribution of $N^+_t$ requires no additional steps
in the MCMC algorithm, other than formally stating that the count data are
``missing'' in future years. An example of this is provided in the Appendix.

\subsection{Bayesian inference}

\citet{dail_madsen:2011} demonstrated how to fit these models using maximum 
likelihood estimation. The likelihood can be maximized numerically using the
\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
are given in the Appendix.

Bayesian inference, however, has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data \citep{link_barker:2010}.    
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. %Although MCMC can be slow, our experience is that it
%may actually converge faster than the amount of time required to
%maximize the likelihood shown in Eq.\ref{eq:like}, especially
%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
%Furthermore, 
MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

Writing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as %\textbf{WinBUGS} and
\textbf{JAGS} \citep{plummer:2003} overcome the technical 
difficulties by allowing users to specify the model using a simple symbolic 
descriptions. Examples are given in the Appendix.

@@

We applied these models to North American Breeding Bird Survey
(BBS) data from 1966-2010 in the bordering US states
Maryland and Virginia. For our focal species, we selected the
ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
forest-breeding migratory songbird with a stable or increasing trend in the region \citep{porneluzi_etal:2011}.
%and golden-winged warblers (\textit{Vermivora chrysoptera}), which
%only breeds in the western parts of these states and has been declining in the
%region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?

@@

We fitted a series of frequentist models, 
and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, 
the previous three with immigration, and exponential + immigration.
%geometric-recruitment +
%immigration, exponential + immigration, Ricker-logistic +
%immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600. % for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with non-informative priors.  We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}. % and, where appropriate, zero-inflated dynamics.  
We tested for lack of convergence using
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.  
We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.

>>>>

The zero-inflated Poisson distribution can be applied to recruitment and 
population growth terms as well as initial abundance. For example, 
the recruitment term of the constant-recruitment DM model
(Eq.~\ref{eq:Nt}) can be modified as follows:
\begin{equation}
G_{i,t} \sim \left\{
\begin{aligned}
\mathrm{Pois}(0) &\; \text{with probability} \; \psi \\
\mathrm{Pois}(\gamma) &\; \text{with probability} \; (1-\psi)\end{aligned} \right.
\label{eq:ZIPts}
\end{equation}
%We have implemented
%zero-inflated dynamics in the Bayesian framework using program
%\textbf{JAGS} \citep[version 3.2.0]{plummer:2003}.
%(version 3.2.0; Plummer 2003)

\subsection{Environmental and demographic stochasticity}
\begin{comment}
Even though the DM model is conceptually simple, %even in
%its most stripped down form 
it includes many random effects (the $N$'s)
making it challenging to compute the likelihood or implement MCMC
algorithms. Nonetheless, in some cases additional random effects
may be of interest. For example, in conventional state-space models,
environmental stochasticity is often modeled as $r_t \sim
\mathrm{Norm}(\mu_r, \sigma_r)$.
\end{comment}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- through %random 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single probability distribution for abundance: $N_{i,t} \sim 
\mathrm{Pois}(f(N_{i,t-1}))$.  
% JEFF: I'm not sure what you're driving at here. Do these papers
% assume a Poisson distribution too?
%This is also a commonly used (and mathematically equivalent) 
%method of modeling demographic stochasticity 
%\citep{bonsall_hastings:2004,melbourne_hastings:2008}.

@@

% JEFF: It seems to me that this approach could be used to account for
% either demographic or enviro stochasticity. How can you know what
% the variation in attributable to?
Another way to model environmental stochasticity is as lognormal variation in observed local population 
growth rate \citep{bjornstad:2001,bonsall_hastings:2004}.  This could be applied to both density-independent and density-dependent models.
For example, for the Ricker-logistic + immigration model:
\begin{equation}
N_{i,t} \sim
\mathrm{Pois}(N_{i,t-1}\exp(\nu_{i,t} + r(1-N_{i,t-1}/K)) + \iota)
\label{eq:nuRand}
\end{equation}
where $\nu$ is a variable with a normal distribution, mean value of 0, and standard deviation $\sigma_\nu$.  
Three variations suggest themselves.  The first has independent environmental stochasticity 
between sites, and might be appropriate where sites are widely dispersed.  In the second, 
all sites have the same environmental conditions in a given year (regional stochasticity, $\nu_{i,t} = \nu_{t}$),
which might be appropriate where the geographic or environmental range is small \citep{hanski:1998}.  
Third, where sites can be close together but the range of sites is broad, a multivariate normal distribution 
for $\nu_{i,t}$ with distance-dependent (geographic or otherwise) correlation coefficients
between sites in a year might be appropriate.

\subsection{Random observer effects}

@@

Current BBS trend estimators deal with this problem by
treating observer identity as a random %(as opposed to a fixed)
effect \citep{link_sauer:2002,sauer_link:2011}.
%(Link and Sauer 2002, Sauer and Link 2011).
%This allows observer-specific differences in detection probability,
%but assumes that observers are selected at random from a pool
%of potential observers.  Models that contain both random and
%fixed effects are referred to as mixed models.  Often in mixed
%models (as in this case), the random effect is modeled not
%because it of interest in itself, but to avoid bias in the
%estimates of the fixed effects.
%
To include random observer effects in DM models, 
% and these extensions, 
Eq.~\ref{eq:p1} can be modified to:
\begin{gather}
X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{N}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
where $X_{i,j,t}$ is the number of individuals counted at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
the standard deviation of the random observer effects (also on the logit scale). 
%We have implemented random observer effects in the Bayesian
%framework using program \textbf{JAGS} \citep[version 3.2.0, see Appendix]{plummer:2003}.
% (version 3.2.0; Plummer 2003)
%with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.

%\subsection{Forecasting Future Population Size}

%State-space models allow for inference about future population size,
%a critical component of population viability analysis (PVA) as well
%as conservation planning. 
%Let $N^+_{t}$ denote population size at some
%future point in time and $[N^+_t]$ denote the probability
%distribution for this random variable. If we can specify this
%distribution, we can make statements about the
%probability that $N^+_t = k$ where $k$ is any possible
%population size including zero.

@@

%Relative to the computational challenges associated with the
%empirical Bayes methods, 
%A fully Bayesian approach to PVA is straightforward to
%implement with MCMC. Because population size at any future point %Furthermore, b
%in time is regarded as a standard, unknown parameter, computing the
%posterior distribution of $N^+_t$ requires no additional steps
%in the MCMC algorithm, other than formally stating that the count data are
%``missing'' in future years. An example of this is provided in the Appendix.

\subsection{Statistical inference}

\citet{dail_madsen:2011} used likelihood based methods for estimating
the parameters of their model. The same likelihood functions can be
used to accomodate the alternative population growth models that we
have proposed. Methods for implementing these models using maximum 
likelihood estimation are included in the %. The likelihood can be maximized numerically using the
\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
are given in the Appendix.

Bayesian inference is an alternative to classical inference with
several appealing features. First, it allows direct probability
statements to be made about a hypothesis given data
\citep{link_barker:2010}.  Second, Bayesian methods offer
straight-forward approaches for combining data from multiple sources
or use existing estimates of parameters as prior distributions. Third,
one common usage of state-space models is for predicting future
population size, and this task easily accomplished using Bayesian
methods.

In
practice, simulation methods such as Markov chain Monte Carlo (MCMC)
are used to simulate the posterior distributions. While time
consuming, MCMC may be the only viable approach for estimating
parameters, especially for hierarchical models with many random
effects.  
% Although MCMC can be slow, our experience is that it
%may actually converge faster than the amount of time required to
%maximize the likelihood shown in Eq.\ref{eq:like}, especially
%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
%Furthermore, 
%MCMC may be the only possible option for estimating parameters
%using some of the extensions described below, such as allowing for
%additional random effects.
%Writing custom MCMC algorthims is often tedious and foreign to
%ecologists, but 
Software packages such as %\textbf{WinBUGS} and
\textbf{JAGS} \citep{plummer:2003} make these methods readily
available to ecologists, and we have provided examples in the
Appendix. 
%overcome the technical 
%difficulties by allowing users to specify the model using a simple symbolic 
%descriptions. Examples are given in the Appendix.

@@

We applied our model extensions to North American Breeding Bird Survey
(BBS) data collected from 1966-2010 in the bordering US states
Maryland and Virginia. For our focal species, we selected the
ovenbird (\textit{Seiurus aurocapilla}), an abundant and widespread,
forest-breeding migratory songbird with a stable or increasing trend
in the region \citep{porneluzi_etal:2011}. 
%and golden-winged warblers (\textit{Vermivora chrysoptera}), which
%only breeds in the western parts of these states and has been declining in the
%region \citep{confer_etal:2011}.  %MORE INFORMATION ABOUT THE SPECIES?

@@

We fitted a series of models using maximum likelihood, 
and then a series of models using MCMC. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, 
the previous three with immigration, and exponential + immigration.
%geometric-recruitment +
%immigration, exponential + immigration, Ricker-logistic +
%immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
%As above, t
These models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600. % for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework with non-informative priors.  
% JEFF: We need to be specific about the priors
We added random observer 
effects and environmental stochasticity, modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}. % and, where appropriate, zero-inflated dynamics.  
We tested for lack of convergence using
five Markov chains for each model \citep{gelman_rubin:1992}.
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.  
We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.
